Football Championship Ticket Options

### 2018-09-15 paper.Rmd
paper updates

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r importPrep}
pacman::p_load(dplyr, ggplot2)

@@

Since its adoption in the 2014-2015 season, the College Football Playoff (CFP) has given college football fans even more reason to be excited about rankings. No longer did a team need to be in the top 2 of the Bowl Championship Series (BCS) to be a championship contender -- instead, the top 4 teams entered into a playoff. In capitalizing on the intersection between excitement and fandom, the CFP teamed with OptionIt to introduce an RSVP system for championship game tickets, starting in 2018. A fan may by an RSVP (an option) for a team. If that team would make it to the championship game, then the fan holding the option would be able to purchase a ticket at face value. In the end, this is likely to save fans a considerable amount of money as compared to purchasing tickets on the secondary market -- tickets on the secondary market were estimated to be an average of XXXX for the 2018 Championship game. In financial modeling terms, this system could be considered a call option for speculation. When buying an option (the RSVP), the buyer has the right to buy a stock (a ticket) in the future at the strike price (the face value of the ticket).

@@

```{r}
top25InitialPrices
```

@@

```{r}
completeData %>% 
  mutate(Rank = as.numeric(Rank), 
         dateTime = lubridate::ymd_hms(dateTime)) %>% 
  dplyr::filter(name %in% top10Ever) %>% 
  select(dateTime, UL_price, LL_price, name) %>% 
  tidyr::gather(key, value, -name, -dateTime) %>% 
  arrange(name, dateTime) %>% 
  ggplot(., aes(dateTime, as.numeric(value), color = key, group = key)) + 
  geom_path(na.rm = TRUE, linejoin = "mitre") + 
  facet_wrap(~ name) +
  theme_minimal()
```

@@

```{r}
volatile %>% 
  filter(name %in% top25Ever) %>% 
  select(volLL, volUL, name) %>% 
  tidyr::gather(key = key, value = value, -name) %>% 
  ggplot(., aes(name, value, color = key)) + 
  geom_point() +
  theme_minimal()
```

@@

```{r}
changes = allHistory %>% 
  select(name, variable, value) %>% 
  distinct() %>% 
  group_by(name) %>% 
  summarize(n = n()) %>% 
  arrange(desc(n)) %>% 
  as.data.frame() %>% 
  head(25) %>% 
  mutate(name = factor(name, levels = .$name))

ggplot(changes, aes(name, n)) + 
  geom_point() + 
  scale_x_discrete(breaks = changes$name) +
  theme_minimal()
```

@@

We can also plot the values:

```{r}

allHistory %>% 
  filter(allHistory$name %in% top25Ever) %>%
  group_by(name, variable) %>% 
  mutate(obs = 1:n()) %>% 
  ggplot(., aes(obs, value, color = variable)) + 
  geom_path() +
  facet_wrap(~name) +
  theme_minimal()
```

@@

```{r, eval = FALSE}
library(RQuantLib)

AmericanOption(type = "call", 
               underlying = , 
               strike = 475, 
               )
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r importPrep}
pacman::p_load(dplyr, ggplot2, stringr, mgcv, broom)

@@

Since its adoption in the 2014-2015 season, the College Football Playoff (CFP) has given college football fans even more reason to be excited about rankings. No longer did a team need to be in the top 2 of the Bowl Championship Series (BCS) to be a championship contender -- instead, the top 4 teams entered into a playoff. In capitalizing on the intersection between excitement and fandom, the CFP teamed with OptionIt to introduce an RSVP system for championship game tickets, starting in 2018. A fan may by an RSVP (an option) for a team. If that team would make it to the championship game, then the fan holding the option would be able to purchase a ticket at face value. In the end, this is likely to save fans a considerable amount of money as compared to purchasing tickets on the secondary market -- tickets on the secondary market were estimated to be an average of \$2,689 for the 2018 Championship game. In financial modeling terms, this system could be considered a call option for speculation. When buying an option (the RSVP), the buyer has the right to buy a stock (a ticket) in the future at the strike price (the face value of the ticket).

There is certainly a huge amount of speculation built into this -- at present an option for a lower-level seat for Alabama is very nearly \$600. This is a considerable financial risk when purchasing such an option. Coupled with a possible face value around \$800, a fan buying an option for \$600 could easily have \$1,400 wrapped up into a ticket. This is still less money than what would be spent on the secondary market, so the reward is potentially high.

@@

```{r}
top25InitialPrices + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_y_continuous(name = "") +
  scale_x_discrete(name = "")
```

@@

```{r}
completeData %>% 
  mutate(Rank = as.numeric(Rank), 
         dateTime = lubridate::ymd_hms(dateTime)) %>% 
  dplyr::filter(name %in% top10Ever) %>% 
  select(dateTime, UL_price, LL_price, name) %>% 
  tidyr::gather(key, value, -name, -dateTime) %>% 
  arrange(name, dateTime) %>% 
  ggplot(., aes(dateTime, as.numeric(value), color = key, group = key)) + 
  geom_path(na.rm = TRUE, linejoin = "mitre") + 
  facet_wrap(~ name) +
  theme_minimal() +
  scale_x_datetime(name = "") +
  scale_y_continuous(name = "")
```

@@

```{r}
volatile %>% 
  filter(name %in% top25Ever) %>% 
  select(volLL, volUL, name) %>% 
  tidyr::gather(key = key, value = value, -name) %>% 
  ggplot(., aes(name, value, color = key)) + 
  geom_point() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_y_continuous(name = "Volatility")
```

@@

```{r}
changes = allHistory %>% 
  select(name, variable, value) %>% 
  distinct() %>% 
  group_by(name) %>% 
  summarize(Count = n()) %>% 
  arrange(desc(Count)) %>% 
  as.data.frame() %>% 
  head(25) %>% 
  mutate(name = factor(name, levels = .$name))

ggplot(changes, aes(name, Count)) + 
  geom_point() + 
  scale_x_discrete(breaks = changes$name) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_x_discrete(name = "")
```

@@

We can also plot the series for all of those teams which have appeared in the top 25 this season:

```{r}
allHistory %>% 
  filter(allHistory$name %in% top25Ever) %>%
  group_by(name, variable) %>% 
  mutate(obs = 1:n()) %>% 
  ggplot(., aes(obs, value, color = variable)) + 
  geom_path() +
  facet_wrap(~name) +
  theme_minimal() +
  scale_x_continuous(name = "") + 
  scale_y_continuous(name = "")
```

@@

#### Explaining Volatility

```{r}
predictors = completeData %>% 
  select(Mean, n, name, dateTime, masseyRecord) %>% 
  mutate(wins = as.numeric(stringr::str_extract(masseyRecord, "^[0-9]+"))) %>% 
  group_by(name) %>% 
  slice(n())

modDat = left_join(volatile, predictors, by = "name")

modLL = gam(volLL ~ s(Mean) + wins + s(n),
            data = modDat)

# summary(modLL)

knitr::kable(tidy(modLL))

knitr::kable(tidy(modLL, parametric = TRUE))

par(mfrow = c(2, 2))

plot.gam(modLL, all.terms = TRUE)

modUL = gam(volUL ~ s(Mean) + wins + s(n),
            data = modDat)

# summary(modUL)

knitr::kable(tidy(modUL))

knitr::kable(tidy(modUL, parametric = TRUE))

par(mfrow = c(2, 2))

plot.gam(modUL, all.terms = TRUE)
```

#### Explaining Current Prices

```{r}
currentPriceModDat = completeData %>% 
  select(Mean, n, name, dateTime, masseyRecord, UL_price, LL_price) %>% 
  mutate(wins = as.numeric(stringr::str_extract(masseyRecord, "^[0-9]+"))) %>% 
  group_by(name) %>% 
  slice(n())

modLL = gam(LL_price ~ s(Mean) + wins + s(n),
            data = currentPriceModDat)

# summary(modLL)

knitr::kable(tidy(modLL))

knitr::kable(tidy(modLL, parametric = TRUE))

par(mfrow = c(2, 2))

plot.gam(modLL, all.terms = TRUE)

modUL = gam(UL_price ~ s(Mean) + wins + s(n),
            data = currentPriceModDat)

# summary(modUL)

knitr::kable(tidy(modUL))

knitr::kable(tidy(modUL, parametric = TRUE))

par(mfrow = c(2, 2))

plot.gam(modUL, all.terms = TRUE)

```

# Take Away Messages

While we don't have data back to the inception of CFP-RSVP, we do have a pretty broad swath. Given our visualizations and our models, the strategy is clear: buy early and buy towards the top of the projected rankings. There is no such thing as parity within college football; of the possible 16 slots combined over the last 4 years, there have been 10 unique teams in the playoffs (with Alabama making 4 appearances, Clemson making 3 appearances, and Ohio State and Oklahoma each having 2 appearances). We would likely invest early in the teams that are likely to be around when the playoffs start.
